\section{Language of Proof Obligations}
\label{wp-lang}

The work of \textsf{WP} consists in translating \textsf{C} and
\textsf{ACSL} constructs into first order logical formul{\ae}. We
denote by $\cal{L}$ the logic language for constructing proof
obligations.  Shortly, this logical language is made of terms
($t:\mathrm{term}$) and propositions ($P:\mathrm{prop}$) that consist of:
\begin{itemize}
\item Natural, signed, unbounded integer constants and their operations;
\item Natural real numbers and their operations;
\item Arrays (as total maps) and records (tuples with named fields);
\item Abstract (polymorphic) data types;
\item Anonymous function symbols with (optional) algebraic properties;
\item Logical connectors;
\item Universally and existentially quantified variables.
\end{itemize}

Actually, the task of the memory model consists in mapping any
heap \textsf{C}-values at a given program point to some variable or term
in the logical $\cal{L}$ language.

\section{The Hoare Memory Model}
\label{wp-model-logical}

This is the simplest model, inspired by the historical definition of
\emph{Weakest Precondition Calculus} for programs with no pointers. In
such programs, each global and local variable is assigned a
distinct variable in $\cal{L}$.

Consider for instance the statement \lstinline{x++;} where
\lstinline{x} has been declared as an \lstinline{int}. In the
\lstinline{Hoare} memory model, this \textsf{C}-variable will be
assigned to two $\cal{L}$-variables, say $x_1$ before the statement, and
$x_2$ after the statement, with the obvious relation $x_2 = x_1+1$ (if
no overflow occurred).

Of course, this model is not capable of handling memory reads or writes
through pointer values, because there is no way of representing
aliasing.

You select this memory model in the \textsf{WP} plug-in with the option
\texttt{-wp-model Hoare}; the analyzer will complain whenever you
attempt to access memory through pointers with this model.

\section{Memory Models with Pointers}
\label{wp-model-pointers}

Realistic memory models must deal with reads and writes to memory
through pointers. However, there are many ways for modeling the raw
bit stream the heap consists of. All memory models $\cal{M}$ actually
implement a common signature:
\begin{description}
\item[Pointer Type:] $\tau$, generally a pair of a base address and an offset.
\item[Heap Variables:] for each program point, there is a set of
  logical variables to model the heap. For instance, you may have a
  variable for the values at a given address, and another one for the
  allocation table. The heap variables $m_1\ldots m_k$ are
  denoted by $\overline{m}$.
\item[Read Operation:] given the heap variables $\overline{m}$, a
  pointer value $p:\tau$, and some \textsf{C}-type $T$, the
  model will define an operation:
  \[\mathrm{read}_T(\overline{m},p) : \mathrm{term}\]
  that defines the representation in $\cal{L}$ of the value of
  \textsf{C}-type $T$ which is stored at address $p$ in the heap.
\item[Write Operation:] given the heap variables $\overline{m}$ before
  a statement, and their associated heap variables $\overline{m}'$
  after the statement, a pointer value $p:\tau$ and a value $v$ of
  \textsf{C}-type $T$, the model will define a relation:
  \[\mathrm{write}_T(\overline{m},p,v,\overline{m}') : \mathrm{prop}\] that relates the
  heap before and after writing value $v$ at address $p$ in the heap.
\end{description}

Typically, consider the statement \lstinline{(*p)++} where
\lstinline{p} is a \textsf{C}-variable of type \lstinline{(int*)}.
The memory model $\cal{M}$ will assign a unique pointer value
$P:\tau$ to the address of \lstinline{p} in memory. 

Then, it retrieves the actual value of the pointer
$\lstinline{p}$, say $A_p$, by reading a value of type
\lstinline{int*} into the memory variables $\overline{m}$ at address
$P$:
\[ A_p = \mathrm{read}_{\mathtt{int*}}(\overline{m},P) \]

Next, the model retrieves the previous \lstinline{int}-value at
actual address $A_p$, say $V_p$:
\[ V_p = \mathrm{read}_{\mathtt{int}}(\overline{m},A_p) \]

Finally, the model relates the final memory state $\overline{m}'$
with the incremented value $V_p+1$ at address $P$:
\[ \mathrm{write}_{\mathtt{int}}(\overline{m},A_p,V_p+1,\overline{m}') \]

\section{Hoare Variables mixed with Pointers}
\label{wp-model-logicvar}

As illustrated above, a very simple statement is generally translated
by memory models into complex formul{\ae}. However, it is possible in
some situations to mix the Hoare memory model with the other ones.

For instance, assume the address of variable \lstinline{x} is never
taken in the program. Hence, it is not possible to create a pointer
aliased with \lstinline{&x}. It is thus legal to manage the value of
\lstinline{x} with the Hoare memory model, and other values with
another memory-model $\cal{M}$ that deals with pointers.

Common occurrences of such a situation are pointer variables. For
instance, assume \lstinline{p} is a variable of type \lstinline{int*};
it is often the case that the value of \lstinline{p} is used (as in
\lstinline{*p}), but not the address of the variable \lstinline{p}
itself, namely \lstinline{&p}. Then, it is very efficient to manage
the value of \lstinline{p} with the Hoare memory model, and the value
of \lstinline{*p} with a memory model with pointers.

Such an optimization is possible whenever the address of a variable is
never taken in the program. It is activated by default in the
\textsf{WP} plug-in, since it is very effective in practice. You can
nevertheless deactivate it with selector ``\texttt{-wp-model raw}''.

\section{Hoare Variables for Reference Parameters}
\label{wp-model-byreference}

A common programming pattern in \textsf{C} programs is to use pointers
for function arguments passed by reference. For instance, consider the
\lstinline{swap} function below:

\begin{ccode}
void swap(int *a,int *b)
{
  int tmp = *a ;
  *a = *b ;
  *b = tmp ;
}
\end{ccode}

Since neither the address of \lstinline{a} nor the one of \lstinline{b}
are taken, their values can be managed by the Hoare Model as described
in the previous section. But we can do even better. Remark that none of
the pointer values contained in variables \lstinline{a} and
\lstinline{b} is stored in memory. The only occurrences of these
pointer values are in expressions \lstinline{*a} and
\lstinline{*b}. Thus, there can be no alias with these pointer values
elsewhere in memory, provided they are not aliased initially.

Hence, not only can \lstinline{a} and \lstinline{b} be managed by the
Hoare model, but we can also treat \lstinline{(*a)} and \lstinline{(*b)}
expressions as two independent variables of type \lstinline{int} with
the Hoare memory model.

For the callers of the \lstinline{swap} function, we can also take benefit
from such by-reference passing arguments. Typically, consider the
following caller program:

\begin{ccode}
void f(void)
{
  int x=1,y=2 ;
  swap(&x,&y);
}
\end{ccode}

Strictly speaking, this program takes the addresses of \lstinline{x}
and \lstinline{y}. Thus, it would be natural to handle those variables
by a model with pointers. However, \lstinline{swap} will actually
always use \lstinline{*&x} and \lstinline{*&y}, which are respectively
\lstinline{x} and \lstinline{y}.

In such a situation it is then correct to handle those variables
with the Hoare model, and this is a very effective optimization in
practice. Notice however, that in the example above, the optimization
is only correct because \lstinline{x} and \lstinline{y} have disjoint
addresses.

These optimizations can be activated in the \textsf{WP} plug-in with
selector ``\texttt{-wp-model ref}'', and the necessary separation
conditions are generated on-the-fly. This memory model first detects
pointer or array variables that are always passed by reference. The
detected variables are then assigned to the Hoare memory model.

This optimization is not activated by default, since the non-aliasing
assumptions at call sites are sometimes irrelevant.

\section{Mixed models hypotheses}
\label{wp-model-hypotheses}

For the previously presented \textsf{Ref} model, but also for the
\textsf{Typed}, and \textsf{Caveat} models presented later, WP lists
the separation and validity hypotheses associated to the choice that
it make of dispatching each pointer and variable either to the
\lstinline{Hoare} or to the model $\cal{M}$ used for the heap.

Consequently, in addition to user-defined function \lstinline{requires},
WP also assumes, and thus states as \lstinline{requires} for the function
caller, that:

\begin{itemize}
\item \lstinline{Hoare} variables are separated from each other,
\item \lstinline{Hoare} variables are separated from the locations in $\cal{M}$,
\item references are valid memory locations,
\item locations assigned via a pointer are separated from \lstinline{Hoare}
  variables whose address is not taken by the function (including via
  its contract).
\end{itemize}

Furthermore, the function must ensure that:

\begin{itemize}
\item locations assigned via a pointer (including the returned value when
  it is a pointer) are separated from \lstinline{Hoare} variables whose address
  is not taken by the function (including via its contract),
\item pointers assigned by the function (including the returned value when
  it is a pointer) are separated from function parameters and \lstinline{Hoare}
  variables whose address is not taken by the function (including via
  its contract).
\end{itemize}

In order to precisely generate these hypotheses, WP needs precise
\lstinline{assigns} specification. In particular each function under
verification and all its callees needs an \lstinline{assigns} specification.
Furthermore, if the function assigns or returns a pointer, WP needs
a correct \lstinline{\from} specification for those pointers. If the
specification is incomplete, a warning \lstinline{wp:pedantic-assigns} is
triggered. Note that WP does not verify that the \lstinline{\from} is correct.

The hypotheses are displayed when the option
\lstinline{-wp-warn-memory-model} is enable (it is enabled by default).
They can be verified by WP using the experimental option
\lstinline{-wp-check-memory-model}.
